Weighted Model Counting with Conditional Weights

### 2020-07-31 paper/paper.tex
Writing

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage{tikz}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{soul}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}

\newtheorem{theorem}{Theorem}
\newtheorem{observation}{Observation}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\Crefname{property}{Property}{Properties}
\Crefname{condition}{Condition}{Conditions}
\creflabelformat{condition}{#2(#1)#3}

\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{NWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\im}{im}

@@

\tikzset{
  Subset/.style={
    draw=none,
    every to/.append style={
      edge node={node [sloped, allow upside down, auto=false]{$\subset$}}}
  }
}

@@

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Potential criticism may be that this doesn't allow us to use SAT-based
  techniques for probabilistic inference. However, they can still be used for a
  significant part of the encoding.
  \begin{itemize}
  \item Zero-probability weights and one-probability weights can be interpreted
    as logical clauses. This doesn't affect ADDMC but could be useful for other
    solvers.
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item
  Algorithms
  \begin{itemize}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} (rediscovered the
    multiplicativity of BAs in different words) (with optimal settings)
  \item Cachet \cite{DBLP:conf/sat/SangBBKP04}
  \item c2d \cite{DBLP:conf/ecai/Darwiche04}
  \item d4 \cite{DBLP:conf/ijcai/LagniezM17} (closed source, boo!)
  \item miniC2D  \cite{DBLP:conf/ijcai/OztokD15}
  \end{itemize}
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item Intuitively, a measure is just like a probability, except it's in
  $\mathbb{R}_{\ge 0}$ instead of $[0, 1]$.
\end{itemize}

\section{Boolean Algebras and Power Sets}

@@

Let $X$ be a set and let $2^X$ be its power set. We can equivalently interpret
$2^X$ as a Boolean algebra. See \cref{tbl:notation} for a summary of the
differences in terminology and notation. We will use both.

@@

\begin{definition}[Operations on functions]
  Let $A\colon 2^X \to \mathbb{R}_{\ge 0}$ and $B\colon 2^Y \to \mathbb{R}_{\ge
    0}$ be functions, $\alpha \in \mathbb{R}_{\ge 0}$, and $x \in X$. We define
  the following operations:
  \begin{description}
  \item[Addition:] $A+B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is a function
    such that $(A+B)(\tau) = A(\tau \cap X) + B(\tau \cap Y)$ for all $\tau \in
    2^{X \cup Y}$.
  \item[Inverse:] $\overline{A}\colon 2^X \to \mathbb{R}_{\ge 0}$ is a function
    such that $\overline{A}(\tau) = 1 - A(\tau)$ for all $\tau \in 2^X$.
  \item[Multiplication:] $A \cdot B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$
    is a function such that $(A \cdot B)(\tau) = A(\tau \cap X) \cdot B(\tau
    \cap Y)$ for all $\tau \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $\alpha A\colon 2^X \to \mathbb{R}_{\ge 0}$ is a
    function such that $(\alpha A)(\tau) = \alpha \cdot A(\tau)$ for all $\tau
    \in 2^X$.
  \item[Projection:] $\exists_xA\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is a function such that $(\exists_xA)(\tau) = A(\tau) +
    A(\tau \cup \{ x \})$ for all $\tau \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}

@@

\begin{definition}[Special functions]
  \phantom{}
  \begin{itemize}
  \item unit $1\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $1(\emptyset) = 1$.
  \item zero $0\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $0(\emptyset) = 0$.
  \item logical constant $[a]\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$,
    $[a](\emptyset) = 0$, $[a](\{a\}) = 1$.
  \end{itemize}
\end{definition}

@@

\section{Weighted Model Counting as a Measure}

\begin{definition}[Measures and weight functions] \label{def:wmc}
  Let $U$ be a set.
  \begin{itemize}
  \item A \emph{measure} is a function $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$
    such that $M(\bot) = 0$, and $M(a \lor b) = M(a) + M(b)$ for all $a, b \in
    2^{2^U}$ whenever $a \land b = \bot$.
  \item A \emph{weight function} is a function $W\colon 2^U \to \mathbb{R}_{\ge
      0}$.
  \item A weight function is \emph{factored} if $W = \prod_{x \in U} W_x$ for
    some functions $W_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$.
  \item We say that a weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
    \emph{induces} a measure $M_W\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
    \begin{equation} \label{eq:induced_measure}
      M_W(x) = \sum_{\{u\} \le x} W(u).
    \end{equation}
  \item A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \emph{factorable} if there exists a factored weight function $W\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $M$.
  \end{itemize}
\end{definition}

@@

In this formulation, the process of calculating the value of $M_W(x)$ for some
$x \in 2^{2^U}$ with a given definition of $W$ is known as \emph{weighted model
  counting}.

\subsection{Relation to the Classical (Logic-Based) View of WMC}

\todo[inline]{Would it be more accurate to replace the word `model' with
  `interpretation' in most situations?}

\begin{table}
  \caption{An overview of notation for a logic defined over two atoms. The
    elements in both columns are listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table}

Let $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge 0}$
be the \emph{weight function} defined by
\begin{align*}
  w(a) = 0.3, \quad w(\neg a) = 0.7, \quad w(b) = 0.2, \quad w(\neg b) = 0.8.
\end{align*}
Let $\Delta$ be a theory in $\mathcal{L}$ with a sole axiom $a$. Then
$\Delta$ has two models, i.e., $\{ a, b \}$ and $\{ a, \neg b \}$. The
\emph{weighted model count} (WMC) \cite{DBLP:journals/ai/ChaviraD08} of
$\Delta$ is then
\begin{equation} \label{eq:wmc_example}
  \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
\end{equation}

Alternatively, we can define $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
\[
  W_a(\{ a \}) = 0.3, \quad W_a(\emptyset) = 0.7
\]
and $W_b\colon 2^{\{q\}} \to \mathbb{R}_{\ge 0}$ as
\[
  W_b(\{ b \}) = 0.2, \quad W_b(\emptyset) = 0.8.
\]
Let $M$ be the measure on $2^{2^U}$ induced by $W = W_a \cdot W_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\[
  M(a) = W(\{ a, b \}) + W(\{ a \}) = W_a(\{a\})W_b(\{b\}) +
  W_a(\{a\})W_b(\emptyset) = 0.3.
\]

Given the theory $\Delta$, we can also compute the probability of a query $b$ as
\cite{DBLP:conf/aaai/SangBK05}
\[
  \Pr_{\Delta, w}(b) = \frac{\WMC(\Delta \land b)}{\WMC(\Delta)}.
\]
The same thing can be accomplished using the algebraic formulation, with $M$
replacing $\mathrm{WMC}$.

@@

\todo[inline,caption={}]{
  Clarify what is meant by`conditional weight function'.
}

@@

  Alternatively, we can let $b$ depend on $a$ and consider weight functions
  $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $W_b\colon 2^{\{a, b\}} \to
  \mathbb{R}_{\ge 0}$ defined as
  \begin{align*}
    W_a(\{a\}) &= 0.9, \\
    W_a(\emptyset) &= 0.1,
  \end{align*}
  and
  \begin{align*}
    W_b(\{a, b\}) &= 0.8, \\
    W_b(\{a\}) &= 0.2, \\
    W_b(\{b\}) &= 0.7, \\
    W_b(\emptyset) &= 0.3.
  \end{align*}
  One can easily check that with these definitions $W$ indeed induces $M$.
\end{example}

@@

\begin{theorem}
  A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  M(u \land v) = M(u)M(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $W_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $W_x(\{ x \}) = M(x)$ and $W_x(\emptyset) =
  M(\neg x)$. Let $M_W$ be the measure induced by $W = \prod_{x \in U} W_x$. We
  will show that $M = M_W$. First, note that $M_w(\bot) = 0 = M(\bot)$ by
  \cref{def:wmc,prop:measure}. Second, let $a = \bigwedge_{u \in U} a_u$ be an
  atom in $2^{2^U}$ such that $a_u \in \{ u, \neg u \}$ for all $u \in U$. Then
  \[
    M_W(a) = W(\{ a_u \mid u \in U \}) = \prod_{u \in U} W_u(\{a_u\}) = \prod_{u
      \in U} M(a_u) = M \left(\bigwedge_{u \in U} a_u \right) = M(a)
  \]

@@

\section{Previous Work}

\subsection{Using WMC to Perform Inference on Bayesian Networks}

Hitherto, four techniques have been proposed for encoding Bayesian networks into
instances of WMC. We will identify them based on the initials of authors as well
as publications years: \texttt{d02} \cite{DBLP:conf/kr/Darwiche02},
\texttt{sbk05} \cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05}
\cite{DBLP:conf/ijcai/ChaviraD05}, and \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}. Below we friendly summarise the observed
performance differences among them. Sang et al. \cite{DBLP:conf/aaai/SangBK05}
claim that \texttt{sbk05} is a smaller encoding than \texttt{d02} with respect
to both the number of clauses and the number of variables but provide no
experimental comparison. Chavira and Darwiche \cite{DBLP:conf/ijcai/ChaviraD05}
compare \texttt{cd05} with \texttt{d02} by measuring the time it takes to
compile either encoding into an arithmetic circuit (but do not measure inference
time). The more recent encoding \texttt{cd05} always compiles faster and results
in a smaller arithmetic circuit (as measured by the number of edges). In their
subsequent paper, the same authors perform two sets of experiments (that are
relevant to this summary) \cite{DBLP:conf/sat/ChaviraD06}. First, they compile
\texttt{cd05} and \texttt{cd06} encodings into d-DNNF (i.e., deterministic
decomposable negation normal form \cite{DBLP:journals/jancl/Darwiche01}),
measuring both compilation time and numbers of edges in the d-DNNF diagram. The
results are mostly in favour of \texttt{cd06}. Second, they compare the
inference time of \texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04}
with the compile times of \texttt{cd05} and \texttt{cd06}, but only on five
(types of) instances. In these experiments, \texttt{cd06} is always faster than
\texttt{cd05}, while the comparison with \texttt{sbk05} is mixed. Sometimes
\texttt{cd06} is orders of magnitude faster than \texttt{sbk05}, sometimes
slightly slower. The performance difference between \texttt{sbk05} and
\texttt{cd05} is even harder to judge: \texttt{sbk05} is better on three out of
five instances and worse on the remaining two. Based on this description, one
would expect \texttt{cd06} to be faster than both \texttt{cd05} and
\texttt{sbk05}, both of which should be faster than \texttt{d02}. The
experiments in \cref{sec:experiments}, however, strongly disagree with this
prediction, showing that the quality of an encoding depends strongly on the
underlying search algorithm or compilation technique.

\subsection{Algebraic Decision Diagrams and Their Use in Probabilistic
  Inference}

Cover:
\begin{itemize}
\item ADDs provide an efficient way to manipulate functions from BAs/power sets
  \cite{DBLP:journals/fmsd/BaharFGHMPS97}.
\item background reading
  \begin{itemize}
  \item Compiling Bayesian Networks Using Variable Elimination (Chavira and
    Darwiche) \cite{DBLP:conf/ijcai/ChaviraD07}
  \item On the Relationship between Sum-Product Networks and Bayesian Networks
    \cite{DBLP:conf/icml/ZhaoMP15}
  \item Using ROBDDs for Inference in Bayesian Networks with Troubleshooting as
    an Example \cite{DBLP:conf/uai/NielsenWJK00}
  \item SPUDD: Stochastic Planning using Decision Diagrams
    \cite{DBLP:conf/uai/HoeySHB99}
  \end{itemize}
\end{itemize}

@@

\todo[inline]{In line with the names of previous encodings, we shall call this
  encoding \texttt{db20}.}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage{tikz}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{observation}{Observation}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\DeclareMathOperator{\im}{im}

@@

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Introduce
  \begin{itemize}
  \item WMC
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} and ADDs
    \cite{DBLP:journals/fmsd/BaharFGHMPS97}
  \item A Bayesian network is a directed acyclic graph with random variables as
    vertices that defines a probability distribution over them. The full
    probability distribution can be expressed as the product of the conditional
    probability distributions of all random variables. For discrete Bayesian
    networks (and we only consider discrete networks in this paper), each such
    distribution can be represented by a table known as a conditional
    probability table (CPT).
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item Our work is, in spirit, similar to... using ROBDDs for Inference in
  Bayesian Networks with Troubleshooting as an Example
  \cite{DBLP:conf/uai/NielsenWJK00}: this is an interesting approach where
  deterministic parts of a BN are expressed as Boolean functions, i.e.,
  extracting the logical from the probabilistic. Maybe I can make a case that my
  encoding (the textual version) kind of does the same, even though the
  algorithm doesn't.
\end{itemize}

\section{Related Work}

\paragraph{Using WMC to perform inference on Bayesian networks.} Hitherto, four
techniques have been proposed for encoding Bayesian networks into instances of
WMC. We will identify them based on the initials of authors as well as
publications years: \texttt{d02} \cite{DBLP:conf/kr/Darwiche02}, \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05},
and \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06}. Below we friendly summarise
the observed performance differences among them. Sang et al.
\cite{DBLP:conf/aaai/SangBK05} claim that \texttt{sbk05} is a smaller encoding
than \texttt{d02} with respect to both the number of clauses and the number of
variables but provide no experimental comparison. Chavira and Darwiche
\cite{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with \texttt{d02} by
measuring the time it takes to compile either encoding into an arithmetic
circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\cite{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \cite{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly $\mathbb{R}$) \cite{DBLP:journals/fmsd/BaharFGHMPS97}.
They have been used to represent value functions in Markov decision processes
\cite{DBLP:conf/uai/HoeySHB99} and, for Bayesian network inference, to represent
each CPT as an ADD \cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs
and arithmetic circuits into a  single representation
\cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous in this
situation because of their ability to fully exploit context-specific
independence, i.e., observable structure within a CPT that is not inherited from
the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

\section{Boolean Algebras, Power Sets, and Propositional
  Logic} \label{sec:prelims}

@@

\begin{table}
  \caption{Notation for a logic with two atoms. The elements in both columns are
    listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table}

\todo[inline]{Completely rewrite this section. It needs to be very clear and
  introduce all relevant terminology.}
\todo[inline]{Would it be more accurate to replace the word `model' with
  `interpretation' in most situations?}

For any set $X$, we can interpret $2^X$ as either a power set or a Boolean
algebra---see \cref{tbl:notation} for a summary of the differences in
terminology and notation.

@@

We build on the definitions of multiplication and projection in the ADDMC paper
\cite{DBLP:conf/aaai/DudekPV20} and define more operations that can be used to
manipulate functions from Boolean algebras to non-negative reals. All of these
operations have efficient implementations in the CUDD \cite{somenzi1998cudd}
package for manipulating ADDs (among other things) that is used by ADDMC
{\cite{DBLP:conf/aaai/DudekPV20}.

\begin{definition}[Operations on functions]
  Let $A\colon 2^X \to \mathbb{R}_{\ge 0}$ and $B\colon 2^Y \to \mathbb{R}_{\ge
    0}$ be functions, $\alpha \in \mathbb{R}_{\ge 0}$, and $x \in X$. We define
  the following operations:
  \begin{description}
  \item[Addition:] $A+B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is
    such that $(A+B)(\tau) = A(\tau \cap X) + B(\tau \cap Y)$ for all $\tau \in
    2^{X \cup Y}$.
  \item[Multiplication:] $A \cdot B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$
    is such that $(A \cdot B)(\tau) = A(\tau \cap X) \cdot B(\tau
    \cap Y)$ for all $\tau \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $\alpha A\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $(\alpha A)(\tau) = \alpha \cdot A(\tau)$ for all $\tau
    \in 2^X$.
  \item[Complement:] $\overline{A}\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $\overline{A}(\tau) = 1 - A(\tau)$ for all $\tau \in 2^X$.
  \item[Projection:] $\exists_xA\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is such that $(\exists_xA)(\tau) = A(\tau) +
    A(\tau \cup \{ x \})$ for all $\tau \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}

@@

\begin{definition}[Special functions]
  We define several special functions:
  \begin{itemize}
  \item unit $1\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $1(\emptyset) = 1$;
  \item zero $0\colon 2^\emptyset \to \mathbb{R}_{\ge 0}$, $0(\emptyset) = 0$;
  \item and $[a]\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$, $[a](\emptyset) = 0$,
    $[a](\{a\}) = 1$ for any $a$.
  \end{itemize}
\end{definition}

@@

\section{WMC as a Measure}

Let $U$ be a set. A \emph{measure} is a function $M\colon 2^{2^U} \to
\mathbb{R}_{\ge 0}$ such that $M(\bot) = 0$, and $M(a \lor b) = M(a) + M(b)$ for
all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$. A \emph{weight function} is
a function $W\colon 2^U \to \mathbb{R}_{\ge 0}$. A weight function is
\emph{factored} if $W = \prod_{x \in U} W_x$ for some functions $W_x\colon
2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that a weight function
$W\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a measure $M_W\colon 2^{2^U}
\to \mathbb{R}_{\ge 0}$ if
\begin{equation} \label{eq:induced_measure}
  M_W(x) = \sum_{\{u\} \le x} W(u).
\end{equation}
Finally, a measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is \emph{factorable}
if there exists a factored weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
that induces $M$.

@@

In this formulation, the process of calculating the value of $M_W(x)$ for some
$x \in 2^{2^U}$ with a given definition of $W$ is known as WMC.

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a \emph{weight function} defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
\end{equation}
Alternatively, we can define $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$W_a(\{ a \}) = 0.3$, $W_a(\emptyset) = 0.7$ and $W_b\colon 2^{\{q\}} \to
\mathbb{R}_{\ge 0}$ as $W_b(\{ b \}) = 0.2$, $W_b(\emptyset) = 0.8$. Let $M$ be
the measure on $2^{2^U}$ induced by $W = W_a \cdot W_b$. Then, equivalently to
\cref{eq:wmc_example}, we can write
\[
  M(a) = W(\{ a, b \}) + W(\{ a \}) = W_a(\{a\})W_b(\{b\}) +
  W_a(\{a\})W_b(\emptyset) = 0.3.
\]

@@

@@

  Alternatively, we can let $b$ depend on $a$ and consider weight functions
  $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $W_b\colon 2^{\{a, b\}} \to
  \mathbb{R}_{\ge 0}$ defined as $W_a(\{a\}) = 0.9$, $W_a(\emptyset) = 0.1$, and
  $W_b(\{a, b\}) = 0.8$, $W_b(\{a\}) = 0.2$, $W_b(\{b\}) = 0.7$, $W_b(\emptyset)
  = 0.3$. One can easily check that with these definitions $W$ indeed induces
  $M$.
\end{example}

Note that in this case we chose to interpret $W_b$ as $\Pr(B \mid A)$
while---with a different definition of $W_b$ that represents the joint
probability distribution $\Pr(A, B)$---$W_b$ by itself could induce $M$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

@@

\begin{theorem}
  A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  M(u \land v) = M(u)M(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $W_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $W_x(\{ x \}) = M(x)$ and $W_x(\emptyset) =
  M(\neg x)$. Let $M_W$ be the measure induced by $W = \prod_{x \in U} W_x$. We
  will show that $M = M_W$. First, note that $M_w(\bot) = 0 = M(\bot)$ by
  \cref{prop:measure} and the definition of a measure. Second, let $a =
  \bigwedge_{u \in U} a_u$ be an atom in $2^{2^U}$ such that $a_u \in \{ u, \neg
  u \}$ for all $u \in U$. Then
  \[
    M_W(a) = W(\{ a_u \mid u \in U \}) = \prod_{u \in U} W_u(\{a_u\}) = \prod_{u
      \in U} M(a_u) = M \left(\bigwedge_{u \in U} a_u \right) = M(a)
  \]

@@

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables to maintain literal independence. In line with the names of previous
encodings, we shall call this encoding \texttt{db20}.
